---
title: "EDA"
author: "Paul K. Yu"
date: "2024-09-23"
output: pdf_document
---

```{r, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, fig.show = 'hide'}
####Analysis of Covariation#########################
####Michelle Nixon##################################
####April 12, 2021##################################
var.explained.eta <- function(posterior){
  eta.data <- posterior$Eta
  eta.pred <- predict(posterior, response = "LambdaX")
  exp.mat <- matrix(NA, nrow=dim(eta.pred)[3],ncol = 1)
  
  for(i in 1:dim(eta.pred)[3]){
    var_fit = sum(apply(eta.pred[,,i], MARGIN = 1, FUN = "var"))
    var_res = sum(apply(eta.pred[,,i] - eta.data[,,i], MARGIN = 1, FUN = "var"))
    var_total = sum(apply(eta.data[,,i], MARGIN = 1, FUN = "var"))
    exp.mat[i,1] = var_fit/ (var_fit + var_res)
  }
  return(exp.mat)
}

##Loading libraries
##Fido needs to be installed using github
##phyloseq and microbiome need to be installed using BioManager 
library(phyloseq)
library(fido)
library(driver)
library(tidyverse)
library(plyr)
library(dplyr)
library(stringi)
library(stringr)

##Setting the seed
set.seed(2021)

#####Part 1: Reading, Filtering, and Processing Metadata/OTU data#####

###Reading in the metadata
metadata = read.delim(file.path("Data", "BritishData", "All_Metadata_2282021.txt"))

##Filtering metadata to Museum of London only with a value for Date_100
metadata_london = metadata %>%
  filter(str_detect(Museum,"MoL")) %>%
  filter(!is.na(BlackDeath_1346_1353)) 

dim(metadata_london)

###Reading in the OTU data
OTU_data = read.delim(file.path("Data", "BritishData", "AllSamples_20210212_RawAbsolute_AllTaxa.txt"))
dim(OTU_data)

##Light cleaning of the names so the taxanomic information is of the same type for every taxa
otu.names.all = str_sub(OTU_data$X.Datasets, end=-2)
otu.names.all = gsub(";[a-zA-Z /]+ group", "", otu.names.all)
otu.names.all = gsub(";[a-zA-Z /]+ incertae sedis", "", otu.names.all)
otu.names.all = gsub(";[a-zA-Z /]+ complex", "", otu.names.all)
otu.names.all = gsub(";[a-zA-Z /]+ subgroup", "", otu.names.all)
otu.names.all = gsub(";[a-zA-Z /]+ subdivisions", "", otu.names.all)
otu.names.all = gsub(";Polyangiaceae", "", otu.names.all)

otu.names.all = stri_list2matrix(str_split(otu.names.all, ";"), byrow=TRUE)
otu.names.all = as.data.frame(otu.names.all)
names(otu.names.all) = c("root", "cellular", "Kingdom", "Phyla", "Class", "Order", "Family", "Genus", "Species")

##Subsetting to the genus level
OTU_data$X.Datasets = otu.names.all$Genus

OTU_data = OTU_data %>%
  filter(!is.na(X.Datasets)) %>%
  group_by(X.Datasets) %>%
  summarise_each(funs(sum)) %>%
  remove_rownames %>% 
  column_to_rownames(var="X.Datasets") 

###Filtering the OTU data to include the samples in the metadata only
###Filter the OTU data
london.names = paste0("X",metadata_london$X.SampleID)
OTU.london = OTU_data[,colnames(OTU_data) %in% c(london.names,metadata_london$X.SampleID)]

##Now, reverse (filtering out metadata to contain only OTU samples) just in case
metadata_london = metadata_london[paste0("X",metadata_london$X.SampleID)%in% colnames(OTU.london), ]

##Checking dimensions
dim(OTU.london)
dim(metadata_london)

##Now, we need to filter out the OTU data.
##First, a quick plot of the rowSums
OTU.london %>% rowSums() %>% ecdf() %>% base::plot() %>% abline(v=1e6)
###Looks like 40% of the samples have counts of zero

###Use a conservative filtering rule
###Filter out all taxa that don't have a count of at least one in 30% of samples
filtered = rowSums(OTU.london > 1) < .25*ncol(OTU.london)
other.tot = colSums(OTU.london[filtered,])
otu.filtered = rbind(OTU.london[!filtered,], "Other" = other.tot)
dim(otu.filtered)

###Reordering metadata_london to match the OTU table
metadata_london = metadata_london[match(colnames(otu.filtered), paste0("X", metadata_london$X.SampleID)),]

latLongs = read.csv(file.path("Data", "BritishData", "cemetryLocations.csv"))

metadata_london$LateDate = ifelse(is.na(metadata_london$LateDate), stri_list2matrix(str_split(metadata_london$Date_100,"_"))[,2],metadata_london$LateDate)

##Creating the final metadata with the needed variables
metadata_prep = metadata_london %>%
  select(X.SampleID, Date_100, Date_200, Date_300, BlackDeath_PrePost, EarlyDate, LateDate, MedievalPostMedieval, Cemetry, MaxillaMandible, BuccalLingual, SubSupragingival, Tooth, Tooth_Simplified, BlackDeath_1346_1353)

##Merging in the mapDamage data
metadata_MD = read.csv(file.path("Data", "BritishData", "metadata_mapDamage.csv")) %>%
  mutate(X.SampleID = Sample) %>%
  select(X.SampleID, DeltaD_mean, DeltaD_mean_methano, DeltaD_mean_por, DeltaD_mean_strep)

metadata_prep = metadata_prep %>%
  join(metadata_MD, by = "X.SampleID")

```

### Load libraries

```{r}
library(ggplot2)
library(tidyverse)
```

### We want to do an exploratory data analysis on the damage data.

```{r}
head(metadata_prep)
```
### Structure of the data

```{r}
str(metadata_prep)
```

### Summary of the data

```{r}
summary(metadata_prep)
```

### From the [mapDamage website](https://ginolhac.github.io/mapDamage/)
### DeltaD, the cytosine deamination probability in double strand context.
### DeltaS, the cytosine deamination probability in single strand context.

```{r}
DeltaD_mean <- metadata_prep$DeltaD_mean
DeltaD_mean_methano <- metadata_prep$DeltaD_mean_methano
DeltaD_mean_por <- metadata_prep$DeltaD_mean_por
DeltaD_mean_strep <- metadata_prep$DeltaD_mean_strep

# For DeltaD_mean:
# Get NA indices
na_indices <- which(is.na(DeltaD_mean))

# Calculate mean
mean_value <- mean(DeltaD_mean, na.rm = TRUE)

# Impute NA values with the calculated mean
DeltaD_mean[na_indices] <- mean_value

# Recheck NA (should be zero)
sum(is.na(DeltaD_mean))

# For DeltaD_mean_methano:
# Get NA indices
na_indices <- which(is.na(DeltaD_mean_methano))

# Calculate mean
mean_value <- mean(DeltaD_mean_methano, na.rm = TRUE)

# Impute NA values with the calculated mean
DeltaD_mean_methano[na_indices] <- mean_value

# Recheck NA (should be zero)
sum(is.na(DeltaD_mean_methano))

# For DeltaD_mean_por:
# Get NA indices
na_indices <- which(is.na(DeltaD_mean_por))

# Calculate mean
mean_value <- mean(DeltaD_mean_por, na.rm = TRUE)

# Impute NA values with the calculated mean
DeltaD_mean_por[na_indices] <- mean_value

# Recheck NA (should be zero)
sum(is.na(DeltaD_mean_por))

# For DeltaD_mean_strep:
# Get NA indices
na_indices <- which(is.na(DeltaD_mean_strep))

# Calculate mean
mean_value <- mean(DeltaD_mean_strep, na.rm = TRUE)

# Impute NA values with the calculated mean
DeltaD_mean_strep[na_indices] <- mean_value

# Recheck NA (should be zero)
sum(is.na(DeltaD_mean_strep))

```

# Plots for DeltaD_mean

```{r}
index <- which(!is.na(DeltaD_mean))

ggplot(data.frame(DeltaD_mean), aes(x = index, y = DeltaD_mean)) +
    geom_point(color = "blue", size = 3, alpha = 0.7) +
    theme_minimal() +
    ggtitle(paste("Scatter plot of DeltaD_mean"))

ggplot(data.frame(DeltaD_mean), aes(x = "DeltaD_mean", y = DeltaD_mean)) +
    geom_boxplot(fill = "blue", alpha = 0.7, width = 0.2) +
    theme_minimal() +
    ggtitle(paste("Box plot of DeltaD_mean"))

ggplot(data.frame(DeltaD_mean), aes(sample = DeltaD_mean)) +
    geom_qq() +
    geom_qq_line(color = "red") +
    theme_minimal() +
    ggtitle(paste("QQ plot of DeltaD_mean"))

shapiro.test(DeltaD_mean)
```

# Plots for DeltaD_mean_methano

```{r}
index <- which(!is.na(DeltaD_mean_methano))

ggplot(data.frame(DeltaD_mean_methano), aes(x = index, y = DeltaD_mean_methano)) +
    geom_point(color = "blue", size = 3, alpha = 0.7) +
    theme_minimal() +
    ggtitle(paste("Scatter plot of DeltaD_mean_methano"))

ggplot(data.frame(DeltaD_mean), aes(x = "DeltaD_mean_methano", y = DeltaD_mean_methano)) +
    geom_boxplot(fill = "blue", alpha = 0.7, width = 0.2) +
    theme_minimal() +
    ggtitle(paste("Box plot of DeltaD_mean_methano"))

ggplot(data.frame(DeltaD_mean_methano), aes(sample = DeltaD_mean_methano)) +
    geom_qq() +
    geom_qq_line(color = "red") +
    theme_minimal() +
    ggtitle(paste("QQ plot of DeltaD_mean_methano"))

shapiro.test(DeltaD_mean_methano)
```

# Plots for DeltaD_mean_por

```{r}
index <- which(!is.na(DeltaD_mean_por))

ggplot(data.frame(DeltaD_mean_por), aes(x = index, y = DeltaD_mean_por)) +
    geom_point(color = "blue", size = 3, alpha = 0.7) +
    theme_minimal() +
    ggtitle(paste("Scatter plot of DeltaD_mean_por"))

ggplot(data.frame(DeltaD_mean), aes(x = "DeltaD_mean_por", y = DeltaD_mean_por)) +
    geom_boxplot(fill = "blue", alpha = 0.7, width = 0.2) +
    theme_minimal() +
    ggtitle(paste("Box plot of DeltaD_mean_por"))

ggplot(data.frame(DeltaD_mean_por), aes(sample = DeltaD_mean_por)) +
    geom_qq() +
    geom_qq_line(color = "red") +
    theme_minimal() +
    ggtitle(paste("QQ plot of DeltaD_mean_por"))

shapiro.test(DeltaD_mean_por)
```

# Plots for DeltaD_mean_strep

```{r}
index <- which(!is.na(DeltaD_mean_strep))

ggplot(data.frame(DeltaD_mean_strep), aes(x = index, y = DeltaD_mean_strep)) +
    geom_point(color = "blue", size = 3, alpha = 0.7) +
    theme_minimal() +
    ggtitle(paste("Scatter plot of DeltaD_mean_strep"))

ggplot(data.frame(DeltaD_mean), aes(x = "DeltaD_mean_strep", y = DeltaD_mean_strep)) +
    geom_boxplot(fill = "blue", alpha = 0.7, width = 0.2) +
    theme_minimal() +
    ggtitle(paste("Box plot of DeltaD_mean_strep"))

ggplot(data.frame(DeltaD_mean_strep), aes(sample = DeltaD_mean_strep)) +
    geom_qq() +
    geom_qq_line(color = "red") +
    theme_minimal() +
    ggtitle(paste("QQ plot of DeltaD_mean_strep"))

shapiro.test(DeltaD_mean_strep)
```
